On Devon Allen's Disqualification at the 2022 World Track and Field Championships

### 2023-07-19 Manuscript/manuscript.tex
Went over many suggestions

The rank based methods described in Section~\ref{sec:rank} are used
to determine if the reaction times for athletes who had competed at other 
championships differed significantly from the 2022 World Championships.  This is
a two group comparison performed two times: once comparing 2019 World
Championships to 2022 World Championships and can be found in 
section~\ref{sec:results2019vs2022} and secondly comparing 2022 national
championships to the 2022 World Championships and can be found in 
~section\ref{sec:resultsnational}.
\jy{Remind the readers what the analysis was about: two-group comparison with
  group being ....}

\subsubsection{2019 World Track and Field Championships} \label{sec:results2019vs2022}
The p value associated with the two group comparison for 2019 and 2022 was 0.001464.  
\eds{what is the ds method?} \of{Its described in the methods section but I dont
think its no longer needed now that we are no longer using the other grouping
method (I believe it was called something like the rosner method)}
This value is for a one-sided test and show that
the mean reaction time on average in 2019 was larger (slower) than the mean 
reaction time
in 2022 for athletes who competed in both championships at an alpha level of 0.01.
This provides good evidence that athletes who competed in both 2019 and 2022 got
faster during that timespan.  \eds{That is not the only explanation}
\subsubsection{2022 National Championships} \label{sec:resultsnational}

  The ``ds" method produced a z score of
3.6069 and p value of 0.000155. This result is highly significant and shows that
the times at the national stage of competition were significantly higher than
the reaction times at the World Championships even for an alpha level of 0.001.

\eds{also consider moving the following paragraph to discussion??}
As Seiko timed both events, there should not be a significant
difference if the timing equipment in both instances was working correctly.
\eds{you just mentioned in the last subsection that Seiko upgraded its 
technology prior to 2022}
  To put it simply, there is not a reasonable explanation for 
this. It is inconceivable that so many athletes would improve such a significant
amount in such a small amount of time. Six countries are represented in this 
comparison, which drastically decreases the chances of this issue being caused by
one specific country.  The United States athletes' improvement in times was
discussed earlier in the paper, but this data shows that on average there was
\jy{any reference?}
improvement for British, Polish, French, Brazilian, and Spanish athletes.

\section{Discussion}\label{sec:Discussion}

Since 1985 Seiko Holding Corporations has served every World Athletics 
Championships as the official timer \citep{wa2022seiko}.  Since 1985, the 
technology and the ability
to accurately predict measurements: long jump distances, false starts, total 
time, reaction time, etc. have all dramatically improved.  Seiko did not start 
tracking reaction time as an official measurement until 1999 in the men's 110 meter
hurdles and women's 100 meter hurdles. 
Seiko regularly updates their equipment so that they provide cutting edge
technology to the World Track and Field Championships, the highest stakes in the 
world of running outside of the Olympics. Seiko's technology for detecting 
reaction times relies on the pressure that athletes exert on the plate when they 
push off.  Their systems measure the time differential
between when the ``gun" goes off to start the race and when the pressure changes 
\citep{wa2022seiko}.  If an athlete has a reaction time under 0.1 seconds, it 
is deemed a 
false start as it is considered that no athlete can react so quickly 
\citep{Seiko-Timing}.  Thus, a reaction time of for example 0.05, suggests that 
the athlete predicted the gun and it was luck that caused their abnormally high 
time.  At the World Championships level, World Athletics and Seiko want to remove 
that element of luck and thus impose the 0.1 second barrier. In 2013, Seiko 
upgraded their timing equipment for sprinting events (110 meter hurdles falls 
under this category) \citep{wa2013backtage}. Since 1999, the
slowest reaction 
times in the men's 110-meter hurdle were recorded in 2013.  That is not to say
 that the higher 
times were caused by Seiko's equipment, but the two may be related.  It is worth 
noting that prior
to 2022, Seiko again upgraded its technology but for its jump management system.
\eds{I would argue that this paragraph should go in the discussion as there are 
no results presented here}

@@

Devon Allen's disqualification at the World Track and Field Championships was
possibly the result of both a faulty timing equipment and Devon Allen
reacting extremely quickly to the start.  Our analysis showed that
the 2022 World Track and Field Championships were a low outlier compared to
every other year in terms of average reaction time.  Additionally, it was shown
through a gamma mixed effects model that there are significant year and race
effects \eds{remove race effects?} that greatly impact reaction time.

@@

Thus while it seems easy
to conclude that Devon Allen was wrongly disqualified, that may not necessarily
be the case. 
If there are issues with reaction time such as athletes approaching 
the 0.1 
second barrier again at the next World Championships (August 2023), then 
World Athletics 
needs to adjust that threshold and allow for faster reaction times so that 
athletes with superb 
reflexes are not penalized. %like how Allen was in 2022.  
World Athletics did not
do anything in response to the \citet{komi2009iaaf} study, but perhaps they 
should now so
that the story of Devon Allen is not repeated.

>>>>

The rank based methods described in Section~\ref{sec:rank} are used
to determine if the reaction times for athletes who had competed at other 
championships differed significantly from the 2022 World Championships.  This is
a two group comparison performed two times: once comparing 2019 World
Championships to 2022 World Championships and can be found in 
section~\ref{sec:results2019vs2022} and secondly comparing 2022 national
championships to the 2022 World Championships and can be found in 
~section\ref{sec:resultsnational}.
\jy{Remind the readers what the analysis was about: two-group comparison with
  group being ....}
\of{Better?}

\subsubsection{2019 World Track and Field Championships} \label{sec:results2019vs2022}
The p value from the two group comparison for 2019 and 2022 was 0.001464.  This 
value is for a one-sided test and show that the mean reaction time on average in
2019 was larger (slower) than the mean reaction time in 2022 for athletes who 
competed in both championships at an $\alpha$ level of 0.01. This provides good
evidence that on average, the reaction times from the 2019 and 2022 World 
Championships are significantly different.  
\eds{That is not the only explanation} \of{You are correct so I changed the 
wording}

\subsubsection{2022 National Championships} \label{sec:resultsnational}
The p value from the two group comparison between the 2022 national
championships and World Championships was 0.000155. This value is for a one sided
test and the result is significant at $\alpha = 0.01$ and shows that
the times at the World Championships were significantly faster than at the
national championships.

\section{Discussion}\label{sec:Discussion}
Since 1985 Seiko Holding Corporations has served every World Athletics 
Championships as the official timer \citep{wa2022seiko}.  Since 1985, the 
technology and the ability to accurately predict measurements: long jump 
distances, false starts, total time, reaction time, etc. have all dramatically 
improved.  Seiko did not start tracking reaction time as an official measurement
until 1999 in the men's 110 meter hurdles and women's 100 meter hurdles. 
Seiko regularly updates their equipment so that they provide cutting edge
technology to the World Track and Field Championships, the highest stakes in the 
world of running outside of the Olympics. Seiko's technology for detecting 
reaction times relies on the pressure that athletes exert on the plate when they 
push off.  Their systems measure the time differential between when the ``gun" 
goes off to start the race and when the pressure changes \citep{wa2022seiko}.  
If an athlete has a reaction time under 0.1 seconds, it is deemed a false start 
as it is considered that no athlete can react so quickly \citep{Seiko-Timing}.  
Thus, a reaction time of for example 0.05, suggests that the athlete predicted 
the gun and it was luck that caused their abnormally high time.  At the World 
Championships level, World Athletics and Seiko want to remove that element of 
luck and thus impose the 0.1 second barrier. In 2013, Seiko upgraded their 
timing equipment for sprinting events (110 meter hurdles falls under this 
category) \citep{wa2013backtage}. Since 1999, the slowest reaction times in the 
men's 110-meter hurdle were recorded in 2013.  That is not to say that the higher 
times were caused by Seiko's equipment, but the two may be related.  It is worth 
noting that prior to 2022, Seiko again upgraded its technology but for its jump 
management system. 
\eds{I would argue that this paragraph should go in the discussion as there are 
no results presented here}
\of{I agree so I moved it here. Not sure if it belongs at the top of the
discussion or elsewhere but here it is.}

\eds{also consider moving the following paragraph to discussion??}
As Seiko timed both events, there should not be a significant
difference if the timing equipment in both instances was working correctly.
\eds{you just mentioned in the last subsection that Seiko upgraded its 
technology prior to 2022}
of{Yes I agree}
To put it simply, there is not a reasonable explanation for 
this. It is inconceivable that so many athletes would improve such a significant
amount in such a small amount of time. Six countries are represented in this 
comparison, which drastically decreases the chances of this issue being caused by
one specific country.  The United States athletes' improvement in times was
discussed earlier in the paper, but this data shows that on average there was
\jy{any reference?}
improvement for British, Polish, French, Brazilian, and Spanish athletes.

@@

Devon Allen's disqualification at the World Track and Field Championships was
possibly the result of both a faulty timing equipment and Devon Allen
reacting extremely quickly to the start.  Our analysis showed that
the 2022 World Track and Field Championships were a low outlier compared to
every other year in terms of average reaction time.  Additionally, it was shown
through a gamma mixed effects model that there are significant year and race
effects \eds{remove race effects?}\of{I think were going to go with the two
effect model so no.} that greatly impact reaction time.

@@

It is confusing as to why the reaction time barrier for men's and women's data
is the same despite numerous articles such as \citep[e.g.,][]{lipps2011implications, 
babicc2009reaction, panoutsakopoulos2020gender} stating differently.  World
Athletics is ignoring differences in gender by setting the reaction time barrier
to be equal. If 0.1 seconds is considered a fair time for women, then 0.1 seconds
could not be considered fair for men, as the probability of observing reaction
times lower than 0.1 would be elevated.

If there are issues with reaction time such as athletes approaching the 0.1 
second barrier again at the next World Championships (August 2023), then World 
Athletics needs to adjust that threshold and allow for faster reaction times so 
that athletes with superb reflexes are not penalized.  World Athletics did not
do anything in response to the \citet{komi2009iaaf} study, but perhaps they 
should now so that the story of Devon Allen is not repeated.
